---
title: "Statistical inference with the GSS data - media exposure by gender"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
library(patchwork)
```

### Load data


```{r load-data}
load("gss.Rdata")

```



* * *

## Part 1: Data


The data is from an observational study, the US General Social Survey, that polls the US population using random stratified sampling method [source:  https://gssdataexplorer.norc.org/documents/441/display].

The data is generalizable only to the US population.

No causality can be established from the data, only an association between variables.

* * *

## Part 2: Research question


**On average, is there a difference in how many hours a day married men and women watch television in the US from year 2000 to 2012?**

This question is of interest to investigate whether in the 21st century there is a difference in media exposure for married men and women. The assumption is that married women might have less media exposure through TV as historically women had more household chores than men thus they might have less time for TV. 
* * *

## Part 3: Exploratory data analysis


* * *
```{r}

#filtering for NA values and year
df1 <- gss %>% 
  filter( !is.na(sex), !is.na(tvhours), year>2000, marital=='Married') 

  
  #plotting
p1<-ggplot(df1, aes(x=sex, y=tvhours)) + 
  geom_boxplot( fill = "#00CCFF") +
  labs(x = "sex", y = "TV hours a day")+
  theme_bw()
p1


info1<-aggregate(tvhours~sex,data=df1,summary)
print(info1)
```
The boxplot and summary statistics show that men and women have the same median and IQR of TV hours a day. 

In this sample, men have on average had 0.07 more TV hours a day than women.

## Part 4: Inference

To test whether there is a difference between the US men and women in the average TV hours a day, we will perform hypothesis testing and calculate confidence interval.





*Hypotheses*


 $H_0: \mu_{TV-hours-men} = \mu_{TV-hours-women}$;
 
 $H_A: \mu_{TV-hours-men} \ne \mu_{TV-hours-women}$ 


*Checking conditions*

* Independence:

  + Within groups 

    - Random sampling is used; 
                
    - Sample size for women is 1921 and men 1699, which is less than 10% of population of married women and men in the US
                 
  + between the groups 
   
    - random sampling is used


* Sample size and skew: 

  + The sample distribution for both men and women is right skewed since the minimum amount of hours a day that a person can watch TV is 0 and a maximum of 24. Taking into account basic human needs as sleeping, the maximum amount of hours decreases to around 18.
  However the sample is large enough thus the sampling distributions of the means is nearly normal. 

```{r}
#examining sample distributions
ggplot(df1, aes(x = tvhours)) +
  geom_histogram(binwidth = .5, colour = "#00CCFF", fill = "#00CCFF") +
  facet_grid(. ~ sex)
```

```{r, message=FALSE}
#examining sampling distributions of the means 
df1men<-df1 %>% 
  filter( sex=='Male')

df1women<-df1 %>% 
  filter( sex=='Female')

sample_means_women_50 <- df1women %>%
                    rep_sample_n(size = 50, reps = 15000, replace = TRUE) %>%
                    summarise(x_bar_women = mean(tvhours))

sample_means_men_50 <- df1men %>%
                    rep_sample_n(size = 50, reps = 15000, replace = TRUE) %>%
                    summarise(x_bar_men = mean(tvhours))



p2<-ggplot(data = sample_means_men_50, aes(x = x_bar_men)) +
  geom_histogram(binwidth = 0.1, colour = "#00CCFF", fill = "#00CCFF")

p3<-ggplot(data = sample_means_women_50, aes(x = x_bar_women)) +
  geom_histogram(binwidth = 0.1, colour = "#00CCFF", fill = "#00CCFF")

p2+p3

```

***Hypothesis testing***

```{r}
inference(y = tvhours, x = sex, data = df1, statistic = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

Hypothesis testing shows that p value is 0.31, which is larger than the significance level of 0.05. Therefore we fail to reject the null hypothesis that there is no difference between the average TV hours a day of married US men and women in the time period 2000-2012.


***Confidence interval***
```{r}
inference(y = tvhours, x = sex, data = df1, statistic = "mean", type = "ci", method = "theoretical")
```

We are 95% confident that the difference in average TV hours a day of married US men and women from 2000 to 2012 is between -0.0654 and 0.2051 hours.

Since 0 is in the confidence interval, we can conclude that there is no difference between married US men and women in the average TV hours a day from 2000 to 2012. This result agrees with the above hypothesis testing method.